Ambient: overview of the workflow and design basics

### 2011-09-18 doc/paper_ambient/paper_ambient.tex
paper introduction + more ref

- Stress Performance of Ambient 
  - in terms of flexible and dynamical resource allocation
  - Stress performance boost derived from this (we need the benchmarks for that!)

@@

As soon as Ambient encounters the flush signal [MORE EXPLANATION]- kernel
execution is initiated along the following sequence:

\begin{itemize}
\item Execution of logistic kernel: allocation of device (MPI group) according to specified request. 
\item Execution of logistic kernel: argument's work groups decomposition according to workDim [INSERT OPENCL REF HERE].
\item Execution of logistic kernel: selected argument's work groups mapping onto device (MPI processes).
\item Transfer of work groups to the target processes.
\item Execution of computational kernel: MPI process executes kernel thread for work items of assigned work groups.
\end{itemize}

>>>>

In the last twenties years, the High Performance Computing (HPC) community has lived a series of technical evolutions that have always influenced the  programming model.
These new paradigms are always introduced by the new products deliver from the computer companies.  If we examined the two last decades, the most powerful
machines have moved from shared memory architecture, powerful but expensive and not  configurable, easily,  to distributed memory with multicores processors using
optional accelerators \cite{Top500}.  This dual transition  material/programming models  were done in three distinctive steps.

Firstly, the introduction of modular computer, based  on independent blade system in the nineties, was a  revolution\footnote{The TOP 500 web site inventories the list of
 the most powerfull computer in the world. It delivers statistic among the architecture; we find the SMP system disappeared of the list in 2002} . It cuts the price of HPC system, and change the programming  model.
 We moved from shared memory API to distributed API. Indeed, as the systems  became independents \df the address space is not shared among nodes, 
  parallel processes have to transmit data over an interconnecting network in order to access
data that other processes have updated. Several technologies were designed to allow communication between nodes (PVM, NX, Express, P4, \etc) \cite{Aoyama:1999}, one in particularly were developed as 
a standard; the Message Passing Interface (MPI) \cite{Snir:1998:MCR:552013}. This interface has been developed by the participation of more 40 organizations, started working on the standard in 1992,
 delivered in 1994 for the first version and 1997 for  the second version. Nowadays, it is the standard for  large grain scalability, it is available under several distributions  from the academic world and HPC constructors.

In the last decade, the distributed memory were confirmed like the technology in HPC architecture machine, 100 \% of the top 500 are now based on this model, as the large scalability applications
used  a main MPI layer, we can cite as popular   in a few scientist field :  [GIVE ME A MPI CODE INTO ALPS !!!!]   {\sc Quantum ESPRESSO} \cite{QE-2009} in quantum chemistry, Abinit \cite{ABINIT-2005, ABINIT-2009} in 
quantum mechanic,   ScaLAPACK \cite{SCALAPACK-1997} in linear  algebra or SMILE in hypersonic rarefied flow \cite{SMILE-2005}. Of course, this short list is not exhaustive, and it can be fill up easily in every fields of 
science and engineering. The limitation of this programming model (the scalability) will be the communication overhead and the workload imbalance among process.

Secondly, the multicore technology  introduced by the ship makers the last decade.  Indeed, the progress of the industrial process allowed the integration of several cores into the same die. 
This technology is now available on all  processor architecture (Power, X86, ARM, \etc). One more time, the programming 
model was modified, and it re-actualized the shared memory model.  To program shared memory machine, a proved technology exists, it is the POSIX threads library. 
The programmers have two possibilities. They utilize the library directly (pThreads) or he lets the compiler generates a multi-thread executables.  
Generally, the former option places the burden on the programmer, but when well done, it provides good performance.
On the other hand,  the second option more knew under the  name OpenMP  \cite{openmp:2002}, easier,  is based on  preprocessor directive. This second option
is clearly the most popular nowadays.

As MPI, OpenMP is developed into a consortium, the version 1.0 was released in 1997, the 2.0 in 2005 and new features were integrated in 2008.
This programming model becomes popular, it is now integrated natively in all compilers and it allows hybrid programs. Hybrid mode MPI/OpenMP
is the actual programming model in HPC \cite{EXASCALE-2010}.  Per example the performance of the ScaLAPACK library can be improved by a multithread BLAS library (MKL, ESSLSMP or ACML). 
Although this ideology gives results, the threads management  gives an additional workload, when the  grain problem size will become too fine \cite{SMITH-2001}.

And then, the present evolution comes from the introduction of the graphic card into HPC, graphic card exists since the eighties \cite{iSBX_1983}, the graphic card market 
was disconnected to the HPC  until a few years ago due to the incredible power of calculation of last generation of GPU. The introducing of the CUDA \cite{CUDA2011} \footnote{An  Open Source alternative to CUDA exists
it is OpenCL, it also extents the CUDA ideology to  multicore processor.}  API allows a real opportunity to exploit graphic card into a cluster.  To summarize the GPU ideology : it is the ability to performed thousand of threads simultaneously, where the threads applied a transformation (kernel)
to a single or a piece of data.

The classical new programming model introduced by the GPU was an hybrid mode were operation are executed  asynchronously between  multiple CPU's core and GPU.
As the pThreads with OpenMP, the programmer can manage by hand or delegate by preprocessor directives  \cite{PGICUDAFOR2010}.
The best example of this programming model is the mathematical library MAGMA \cite{MAGMA_2011}.

In this paper, we proposed an alternative approach of programming.  We developed a powerful oriented object framework named Ambient, that transposed  the CUDA ideology to a cluster. 
As in the CUDA framework, in Ambient, the developer writes kernels which are executed in parallel, dynamically, and independently if needed, all other the resource of the cluster.
The needed communications  are assured by a nested MPI layer away from the users. From the user point of view, the objective is to write  parallel program without interacting the MPI API.

In this paper, we are going to describe the difference of workflow between an Ambient  and a CUDA program,  describe the blue print of Ambient and evaluated on several  benchmarks.  
Until the end of this paper, we will apply and illustrate our framework on basic linear algebra problems.

@@

As soon as Ambient encounters the flush signal (contrary to CUDA, the programmer orders the execution   as in queue model) kernel
execution is initiated along the following sequence:

\begin{itemize}
\item Execution of logistic kernel: allocation of device (MPI group) according to specified request. 
\item Execution of logistic kernel: argument's work groups decomposition according to workDim \cite{OPENCL2011}.
\item Execution of logistic kernel: selected argument's work groups mapping onto device (MPI processes).
\item Transfer of work groups to the target processes.
\item Execution of computational kernel: MPI process executes kernel thread for work items of assigned work groups.
\end{itemize}
